---
title: "StatSim & CM - Exercise 1 Stadler"

output:
  pdf_document:
    toc: yes
    toc_depth: 4
    number_sections: no
    df_print: paged
date: "2024-10-07"
---

```{r setup, include=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


# Exercise 1 - Introduction to Simulation with Variance Estimation

## Compare the 4 algorithms against R's 'var' function as a gold standard regarding the quality of their estimates.

### Implement all variants of variance calculation as functions.
Code for algorithm 1-4
```{r algo-1to4, warning=FALSE , message=FALSE}
var_1 <- function(){
  value<-0
  return(value)
}
```



### Write a wrapper function which calls the different variants.
Code for wrapper
```{r wrapper, warning=FALSE , message=FALSE}

```

## Compare the computational performance of the 4 algorithms against R's 'var' function as a gold standard and summarise them in tables and graphically.



## Investigate the scale invariance property for different values and argue why the mean is performing best as mentioned with the condition number.

### Compare the results according to the instructions provided by Comparison I and Comparison II of the slides.

### Provide your results in table format and graphical format comparable to the example in the slides.

## Compare condition numbers for the 2 simulated data sets and a third one where the requirement is not fulfilled, as described during the lecture.